---
title: "Box plot and time series combo"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Since people were curious, I needed to write down how I did it, and I appreciated an excuse to mess with R Markdown, the gist of how I overlayed individual time series onto group box plots is below. The `.rmd` file is also attached.

We've taken to visualizing any changes during our experiments using boxplots that show group behavior before, during, and after the experiment. However, this assumes that the observations are independent, which they arent: we have multiple observations for each individual which correlated through time. I wanted to be able to see what the time series for each individual was within the box I've just put them in. So, here's how I was able to overlay individual time series on a box and whisker plot.

```{r}
library(ggplot2); library(dplyr)

my_data <- tibble(
  individual = rep(letters[1:5], each = 15),
  time = rep(
    seq.POSIXt(ISOdate(2020, 02, 03), by = 'min', length.out = 15),
    times = 5),
  phase = rep(c('Pretrial', 'Trial', 'Posttrial'), each = 5, times = 5),
  value = sample(1:10, 75, replace = T)
)

my_data

```

Find start/end time for each trial phase and join back in so that we can eventually find where the observation falls in that phase's timeline.

```{r}
my_data <- my_data %>% 
  group_by(phase) %>% 
  summarize(time.start = min(time),
            time.end = max(time)) %>% 
  right_join(my_data)

my_data

```

POSIX dates are really just the number of seconds since midnight on Jan 1, 1970.

```{r}
ISOdate(1970, 01, 01, 00, 00, 00, tz = 'UTC')
as.numeric(
  ISOdate(1970, 01, 01, 00, 00, 00, tz = 'UTC')
)
as.POSIXct(0, origin = '1970-01-01', tz = 'UTC')


ISOdate(1970, 01, 01, 00, 00, 01, tz = 'UTC')
as.numeric(
  ISOdate(1970, 01, 01, 00, 00, 01, tz = 'UTC')
)
as.POSIXct(1, origin = '1970-01-01', tz = 'UTC')
```

We can't do math on a date, but we can use the number of seconds. So, convert the POSIX dates to numeric. We will use this to translate the time line to a different scale in the next step, below.

```{r}
my_data <- my_data %>% 
  mutate_at(vars(starts_with('time')),
            as.numeric)

my_data
```

So... what does a boxplot of this data look like?

```{r}
ggplot(data = my_data, aes(x = phase, y = value)) +
  geom_boxplot()
```

What if we overlay the raw values?

```{r}
ggplot(data = my_data, aes(x = phase, y = value)) +
  geom_boxplot() +
  geom_point(aes(color = individual))
```

We can see that all of points fall on a line in the dead center of the boxes. `ggplot` thinks of the total area available to each box as equal to 1, running from -0.5 to 0.5, with 0 being the center (where the box whiskers are). This includes the space between boxes, so ~2/3 of the space given for the box and 1/3 is given to space on either side.

We want to translate the observation time (within the given trial phase start/end range) to a range that can fit within the boxes of the box plot. Since 2/3 of the space provided for each box is given to the box, itself, this range is from -1/3 to +1/3. We will use this number to "nudge" the observation to the left (negative values) or right (positive values) of the center of the box.

```{r}
my_data <- my_data %>% 
  mutate(nudge = (((time - time.start) * (1/3 - (-1/3))) /
                    (time.end - time.start)) - 1/3)

my_data
```

How do we "nudge" the values? In the `position` argument of `geom_point`, we can provide how much we want the values "nudged" using the `position_nudge` function in `ggplot`. Note that this function is not aware of what data you're using, so you have to give the full reference using `$`.

```{r}
ggplot(data = my_data, aes(x = phase, y = value)) +
  geom_boxplot() +
  geom_point(aes(color = individual),
             position = position_nudge(x = my_data$nudge))
```

The points are where they're supposed to be, now we just have to connect them!

```{r}
ggplot(data = my_data, aes(x = phase, y = value)) +
  geom_boxplot() +
  geom_point(aes(color = individual),
             position = position_nudge(x = my_data$nudge)) +
  geom_line(aes(color = individual),
            position = position_nudge(x = my_data$nudge))
```

