---
title: "FOR loops"
output: html_notebook
---

FOR loops are a pretty basic part of programming; you tell the computer how many times you want something done, then tell the computer what it is that you want it to do. Basically:

* *FOR* each part of a sequence, run a function

Parts of a FOR loop:

* Call to the `for` function
* An index that will change every iteration (usually `i` is used)
* A sequence of data (numbers, letters, anything really)
    * Each part of the data will be assigned to the index in turn
    * `seq(1, 10, 1)`
    * `list(seq(1, 10, 1), seq(1, 100, 10), seq(1, 1000, 100))`
    * `data.frame(a = seq(1, 10, 1), b = seq(1, 100, 10), c = seq(1, 1000, 100))`
* A function
    * `function(x){x + 2}`
    
```{r}
for(i in seq(1, 10, 1)){
  print(i + 2)
}
```

What's actually going on here? Here's what the data look like:

```{r}
seq(1, 10, 1)
```

The FOR loop first takes the first element of our sequence and assigns it to i.
```{r}
i <- 1
```
Now, every `i` in the loop is set equal to 1. `print(i + 1)` becomes `print(1 + 1)` and this is evaluated.

```{r}
print(1 + 1)
```

The first iteration is now complete. The loop will run *FOR* as many iterations as needed. Here, there are 10.

    i <- 2
    print(2 + 2)
    
Next iteration

    i <- 3
    print (3 + 2)

And so on...

##Use `i` as an index
`i` doesn't have to represent your data directly: it can be used as an index. It
also doesn't need to be called `i`! Let's use a FOR loop to calculate the sum of
each column of a data frame, using the word `column` to represent the column index.

```{r}
data <- data.frame(a = seq(1, 100, 10), b = seq(1, 10, 1), c = seq(1, 1000, 100))
column_sums <- NULL

for(column in seq(1, 3, 1)){
  column_sums[column] <- sum(data[, column])
}

column_sums
```

This is useful when we get to writing functions--it means that your loops can be generalized.
```{r}
data <- data.frame(a = seq(1, 100, 10), b = seq(1, 10, 1),
                   c = seq(1, 1000, 100), d = seq(1, 10000, 1000))

for(column in seq(1, ncol(data), 1)){
  column_sums[column] <- sum(data[, column])
}
column_sums
```

It's also useful for when the output of a function depends on a previous calculated value.
```{r}
data <- rep(NA, 10)
data[1] <- 1
data
```
```{r}
for(i in seq(1, length(data), 1)){
  data[i + 1] <- data[i] + 2
}

data
```

##The `apply` family
Whereas FOR loops are basic programming, the `apply` functions are vectorized loops in R. For a plain-English explanation of what that means, check out [this blog post](http://www.noamross.net/blog/2014/4/16/vectorization-in-r--why.html). If we wanted to use `apply` to calculate the column sums as above, we need to supply:

* The data
* The margin over which we want to apply the function (columns are the second margin)
* The function we wish to apply

```{r}
data <- data.frame(a = seq(1, 100, 10), b = seq(1, 10, 1), c = seq(1, 1000, 100))
apply(data, 2, sum)
```
Rows are the first margin (for comparison's sake). Here are the row sums:
```{r}
apply(data, 1, sum)
```


We can also `apply` custom functions:
```{r}
apply(data, 1, function(x) x + 2)
apply(data, 2, function(x) x + 2)
```

###`lapply`
`lapply` is a special version of `apply` that creates a list. You don't need to specify the margin with `lapply`, as it assumes it will be applying something to each of the list's elements. This is the function in the `apply` family I use the most, as it lets me apply a function simultaneously to different datasets that may or may not be the same size.

```{r}
data <- data.frame(a = runif(100), b = rnorm(100),
                   c = rexp(100),
                   id = c(rep('kitty', 40), rep('dog', 50), rep('mouse', 10)))

# Split the data frame into a list according to id. These chunks will be 
# different sizes.
data <- split(data, data$id)
```

I want to see the first few rows of each element of `data`, but using `head(data)` returns everything! 

```{r}
head(data)
```


This is because `head` is asking `R` to return the first few elements of the input; if it were a data frame, it'd be returning 10 rows, but since it's a list, it returns 10 elements. What I actually need to do is apply `head` to each element of the list independently. This is where `lapply` starts to make an appearance.

```{r}
lapply(data, head)
```

What if we want to apply the same model to each of the data sets? It can now be done quickly and with one line of code...

```{r}
model <- lapply(data, function(x) lm(x, formula = c ~ a))
model
```

Model coefficients and residuals? Also quick and easy.

```{r}
lapply(model, coef)
lapply(model, residuals)
```

##Added bonus to learning this: **Parallelization**
A lot of problems like the above are referrred to as ["embarassingly parallel"](http://en.wikipedia.org/wiki/Embarrassingly_parallel), where each task is completely separate from the others. This means that you can, theoretically, send each of the above models to different computers and have them work independently. That's a bit hard, but if you have a lot of data, you can send each problem to a different *part* of your computer that may or may not have been utilized before, then gather the information when everything is done running. This is **paralellization** and can really speed up your analysis after you master `lapply`. Topic for another day, but if you want to dig in, [this is a good starting point](http://gforge.se/2015/02/how-to-go-parallel-in-r-basics-tips/).