---
title: "VR2C Communication and Output with R"
output:
  html_document:
    df_print: paged
---

After installing the [SeaCOM drivers](https://www.sealevel.com/support/category/driver-downloads-software/) as outlined in the [VR2C Users Manual](https://vemco.com/wp-content/uploads/2014/03/vr2c-manual.pdf), we can use the [`serial`](https://cran.r-project.org/web/packages/serial/serial.pdf) package in R to communicate with the VR2C.

First, construct the serial interface connection using the com port number that is connected to the VR2C ("COM3" here) and a character string with the VR2C's configured baud rate (9600 here), parity ("n" for "none"), data bits (8) and stop bits (1). `translation` should be `cr`, `crlf`, or `auto`. Then open the connection.

```{r}
library(serial)

vr2c_con <- serialConnection('vr2c_1', port = 'COM3',
                             mode = '9600,n,8,1',
                             buffering = 'line',
                             newline = 1,
                             handshake = 'none',
                             translation = 'crlf')
open(vr2c_con)
```

Wake up the receiver using any ASCII character, give the receiver time to wake up, and issue a command using the command strings outlined in the manual. Note that you need to start the command with an asterisk.

```{r}
write.serialConnection(vr2c_con, '')
Sys.sleep(0.1)
write.serialConnection(vr2c_con, '*450280.0#19,status')
```

Read the response.

```{r}
response <- NULL
now <- Sys.time()

while(Sys.time() < now + 1){
  response <- c(response, read.serialConnection(vr2c_con))
}

response
```

If you get no response, try re-sending a ASCII character to wake it up. Start the receiver.
```{r}
write.serialConnection(vr2c_con, '*450280.0#19,rtmnow')
Sys.sleep(0.1)
read.serialConnection(vr2c_con)
```

Read in detections.
```{r}
write.serialConnection(vr2c_con, '')
write.serialConnection(vr2c_con, '*450280.0#19,rtmnow')

response <- NULL
now <- Sys.time()

while(Sys.time() < now + 1){
  response <- c(response, read.serialConnection(vr2c_con))
}

response
```


Close the connection.

```{r}
close(vr2c_con)
```

