---
title: "Mackerel Random Forest code"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Mackerel classification through Random Forest
This function utilizes the randomForest R package (Liaw and Wiener 2002). Data are those collected in Secor 2017; values are country, subregion, and year of capture, as well as calculated age, d18O, and d13C values.

```{r}
library(randomForest)

mack.data <- read.csv('scripts/mackerel/alldata2.csv', na.strings = 'n/a')
head(mack.data)
```

Separate random forest models are built using age-1 fish from the 1998, 1999, or 2000 year class.

```{r}
baseline_1998 <- mack.data[mack.data$Yearclass == 1998 & mack.data$Age == 1,]
baseline_1998$Country <- droplevels(baseline_1998$Country)

rf_1998 <- randomForest(Country ~ d18O, data = baseline_1998, ntree = 10000,
                           importance = TRUE, proximity = TRUE, oob.prox = T)
rf_1998
```

Classification error varies according to origin and year class. Using the trained random forest models, origin of older fish can be predicted. Here, we classify age-3 fish from the 1998 year class.

```{r}
to.classify <- mack.data[mack.data$Age == 3 &
                           mack.data$Yearclass == 1998,]

# Nominal classification (>50%)
classification <- predict(rf_1998, newdata = to.classify, type = "response")

# Probability 
prob <- predict(rf_1998, newdata = to.classify, type = "prob")

head(cbind(to.classify, classification, prob))
```

The operations can be combined into a single function.

```{r}
class.func <- function(data, year_class, age, country = 'US'){
  # Build random forest model using baseline data (age-1 fish of a given year class)
  baseline <- data[data$Yearclass == year_class & data$Age == 1,]
  baseline$Country <- droplevels(baseline$Country)
  
  #Train the Random Forest model
  rf.model <- randomForest(Country ~ d18O + d13C, data = baseline, ntree = 10000,
                           importance = TRUE, proximity = TRUE, oob.prox = T)
  
  # Use trained random forest model to classify age/yearclass/country subset
  to.classify <- data[data$Age == age &
                        data$Yearclass == year_class &
                        data$Country %in% country,]
  
  # Nominal classification (>50%)
  Classification <- predict(rf.model, newdata = to.classify, type = "response")
  
  # Probability 
  prob <- predict(rf.model, newdata = to.classify, type = "prob")
  
  cbind(to.classify, Classification, prob)
}

age3_1998 <- class.func(mack.data, 1998, 3)
head(age3_1998)
```

This allows multiple ages and year-classes to be analyzed simultaneously.


```{r}
class <- matrix(c(2, 1998,
                  2, 1999,
                  2, 2000,
                  3, 1998,
                  3, 2000,
                  4, 1999,
                  5, 1998), ncol = 2, byrow = T)
colnames(class) <- c('age', 'year_class')

all.data <- apply(class, MARGIN = 1,
                  function(x) class.func(mack.data, year_class = x['year_class'],
                                                   age = x['age']))
all.data <- do.call(rbind, all.data)

head(all.data)
```

